Generating Empirically Optimized Numerical Software from MATLAB Prototypes

### 2008-09-12 doc/papers/cgo08/paper.tex
version with abstract
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@348 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

\usepackage{url}
\usepackage{html}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{amsmath}
\usepackage[margin=1in]{geometry}

@@

\abstract{TODO}

@@

>>>>

\usepackage{url}
\usepackage{html}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{cite}
\usepackage{amsmath}
\usepackage[margin=1in]{geometry}

@@

\abstract{
The growing demand for higher levels of detail and accuracy in results means that the size and complexity of scientific computations is increasing at least as fast as the improvements in processor technology. Programming scientific applications is hard, and optimizing them for high performance is even harder.  The development of optimized codes requires extensive knowledge, not only of the costs of floating-point arithmetic but also of  memory access issues and compiler optimizations. Experiments show that the complexity of this hardware-software system means that performance is difficult to predict fully.  Therefore, computational scientists are often forced to choose between investing too much time in tuning code or  accepting performance that is significantly lower than the best achievable performance on a given architecture.

In this paper, we describe the first steps toward a fully automated system for the optimization of the matrix algebra kernels that are a foundational part of many scientific applications.  To generate highly optimized code from a high-level MATLAB prototype, we will follow a three-step approach.  To begin, we have developed a compiler that converts a MATLAB script into simple C code.   We then use the source transformation tool PLuTo to optimize that code for coarse-grained parallelism and locality simultaneously. The result is an automatically generated code containing loop transformations and OpenMP annotations for exploiting multicore parallelism. Finally, we employ the empirical performance tuning system Orio to generate many tuned versions of the same operation using different optimization techniques, such as loop unrolling and memory alignment. Orio performs an automated empirical search for selecting the best among the multiple optimized code variants.

We discuss performance results showing that the code generated by using our system significantly outperforms not only the original simple C code but also code based on source BLAS, ATLAS-optimized BLAS, and Intel MKL routines.
}

@@

This motivates further work on automating more parts of the process, so that these tools can be used by application developers to automatically produce custom linear algebra libraries tuned for specific architectures.
